Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-30 overleaf/vectorization.tex
Edits to Sec. 5.

There are two key phases of the algorithm. Phase 1 applies dependence analysis and inserts
raise dimension and drop dimension operations as necessary. After this phase operations work
on arrays of the corresponding dimensionality and we optimistically vectorize all arrays.

Phase 2 takes into account dependence analysis. It proceeds from the inner-most towards the outer-most 
loop. For each loop it anchors dependence cycles (closures) around pseudo PHI nodes, removing vectorization 
from the dimension of the loop.

There are several important points in this phase. We may break a loop into smaller loops which would discover opportunities 
for optimization in intermediate statements in the loop. We remove redundant back edges into PHI nodes that are 
due entirely to arrays, which opens up opportunities as well. The pseudo-code in red deals with this extension
of the basic algorithm. We determine that some array pseudo-PHIs are \emph{target-less}, meaning
that there is no loop-carried dependence for that array and back-edges into these nodes can be safely removed. 
This opens up opportunities for vectorization across the write array. Handling of target-less nodes is non-trivial, 
as in certain cases the node cannot be removed. \secref{sec:extension_with_writes} details our algorithm.

\ana{Add an example from histogram here.}

Phases 3 and 4 are less important --- phase 3 cleans up local arrays of references and phase 4 turns operations into MOTION operations.

@@

\begin{figure*}%[!b]
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~~~~~~~~~}l@{~~~~~~~~~}l@{~~~~~~~~~}l}
  s &::= s_1;s_2 & \gamma(s) = \gamma(s_1) \; ; \; \gamma(s_2) & \mathit{sequence} \\
     & \mid \x[i,J,k] = {\sf op\_SIMD}(\y[i,J,k], \z[i,J,k]) & \gamma(\x[i,J,k] = {\sf op\_SIMD}(\y[i,J,k], \z[i,J,k])) = & \mathit{operation} \\ 
     &                                                                               & ~~\x[i,0,k] = \y[i,0,k] \; {\sf op} \; \z[i,0,k] \;\; || &  \\
     &                                                                               & ~~~~\x[i,1,k] = \y[i,1,k] \; {\sf op} \; \z[i,1,k] \;\; || \; ...\; || & \\
     &                                                                               & ~~~~~~\x[i,J\!-\!1,k] = \y[i,J\!-\!1,k] \; {\sf op} \; \z[i,J\!-\!1,k] & \\
     & \mid \x[i,J,k] = {\sf const} & \mbox{analogous} & \mathit{constant} \\
     & \mid \x[i,J,k]  = {\sf PHI}(\x_1[i,J,k],\x_2[i,J,k\!-\!1]) & & \mathit{pseudo} \; \mathit{PHI} \\
     & \mid \x[i,J,k] = {\sf raise\_dim}(\x'[i],(J\!:\!J,k\!:\!K)) & & \mathit{raise} \; \mathit{dimension(s)} \\
     & \mid \x[i,J] = {\sf drop\_dim}(\x'[i,J,k],k) &  & \mathit{drop} \; \mathit{dimension(s)} \\
%     & \mid \x = \y & & \mathit{propagation} \\
     & \mid {\sf for} \; i \; {\sf in} \; {\sf range}(I): s  & \gamma({\sf for} \; i \; {\sf in} \; {\sf range}(I): s) = & \mathit{loop} \\
     & & ~~\gamma(s)[0/i] \; ; \; \gamma(s)[1/i] \; ; \; ... \; ; \; \gamma(s)[I\!-\!1/i] & \\
  \end{array}    \\
\end{array}
\]
\vspace{-3mm}
\caption{MPC Source Syntax and Semantics. $\gamma$ defines the semantics of MPC source which is a linearization of MPC Source. A SIMD operation parallelizes operations across the vectorized $J$ dimension. $||$ denotes parallel execution, which is standard. $\gamma$ of a for loop unrolls the loop. $;$ denotes sequential execution. Iterative MPC Source trivially extends to non-vectorized dimensions over the enclosing loops.}\label{fig:MPC_Source_syntax}
\label{fig:mpc_syntax}
\vspace{-5mm}
\end{figure*}

@@

\figref{fig:mpc_syntax} states the syntax and linearization semantics of MPC Source. For simiplicity,
we consider only scalars and read-only arrays, however, the treatment extends to write arrays as well.
$\x[i,J,k]$ denotes the value of scalar variable \x{} at loop nest $i,j,k$. 
%(We use 3 dimensions, $i,j,k$, however, the treatment generalizes to arbitrary dimensions.) 
Upper case $J$ denotes a vectorized dimension and lower case $i,k$ denote iterative dimensions. Our compiler imposes 
semantic restrictions over the syntax: (1) $\x$ is treated as a 3-dimensional array and (2) $\x[i,J,k]$ must be enclosed
into for-loops on non-vectorized dimensions $i$ and $k$:
\begin{pythonn}
for i in range(I):
   ...
   for k in range(K):
      ... x[i,J,k] ...
\end{pythonn}

@@

For each MPC Source program $a$ we compute the def-use edges in the standard way: if base statement $s1 \in a$ defines variable \x, e.g., $\x[i,j,k] = ...$, 
and base statement $s2 \in a$ uses \x, e.g., $... = ... \x[i,j,k]$ and there is a path in CFG from $s1$ to $s2$, then there is a def-use edge from
 $s1$ to $s2$. We extend the dimensionality of a statement into $s1[i,j,k]$ where $s1[i,j,k]$ inherits the dimensionality of the left-hand-side of the assignment.

 \ana{Add on the linearization semantics?}

@@

Array writes limit vectorization as they sometimes introduce infeasible loop-carried dependencies. Consider the following example: \ana{Have to add citation to Aiken's paper}

>>>>

There are two key phases of the algorithm. Phase 1 inserts raise dimension and drop dimension 
operations according to def-uses. E.g., if there is an inner-to-outer dependence, it inserts $\mathit{raise\_dim}$,
and similarly, if there is an outer-to-inner dependence, it inserts $\mathit{drop\_dim}$. After this phase operations work
on arrays of the corresponding dimensionality and we optimistically vectorize all arrays.

Phase 2 proceeds from the inner-most towards the outer-most loop. For each loop it anchors dependence cycles (closures) 
around pseudo PHI nodes then removes vectorization from the dimension of that loop. There are two important points in this 
phase. First, it may break a loop into smaller loops which would discover opportunities 
for vectorization in intermediate statements in the loop. Second, it handles write arrays. 
It creates opportunities for vectorization in the presence of write arrays, even though 
Cytron's SSA adds a back-edge to the array PHI-node, thus killing vectorization of statements
that read and write that array.

The excerpts in {\color{red} red} color in the pseudo code below highlight the extension 
with array writes. We advise the reader to omit the extension and assume just read-only
arrays. We explain the extension in \secref{sec:array_writes} details our algorithm. 
(As many of our benchmarks include array writes, it plays an important role.)

%The pseudo-code in red deals with this extension
%of the basic algorithm. We determine that some array pseudo-PHIs are \emph{target-less}, meaning
%that there is no loop-carried dependence for that array and back-edges into these nodes can be safely removed. 
%This opens up opportunities for vectorization across the write array. Handling of target-less nodes is non-trivial, 
%as in certain cases the node cannot be removed. \secref{sec:extension_with_writes} details our algorithm.

Phases 3 cleans up local arrays of references (this is an optional phase and our current implementation does not include it)
and Phase 4 explicitly turns operations into MOTION SIMD operations.

@@

\begin{figure*}%[!b]
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~~~~~~~~~}l@{~~~~~~~~~}l@{~~~~~~~~~}l}
  s &::= s_1;s_2 & \gamma(s) = \gamma(s_1) \; ; \; \gamma(s_2) & \mathit{sequence} \\
     & \mid \x[i,J,k] = {\sf op\_SIMD}(\y_1[i,J,k], \y_2[i,J,k]) & \gamma(\x[i,J,k] = {\sf op\_SIMD}(\y_1[i,J,k], \y_2[i,J,k])) = & \mathit{operation} \\ 
     &                                                                               & ~~\x[i,0,k] = \y_1[i,0,k] \; {\sf op} \; \y_2[i,0,k] \;\; || &  \\
     &                                                                               & ~~~~\x[i,1,k] = \y_1[i,1,k] \; {\sf op} \; \y_2[i,1,k] \;\; || \; ...\; || & \\
     &                                                                               & ~~~~~~\x[i,J\!-\!1,k] = \y_1[i,J\!-\!1,k] \; {\sf op} \; \y_2[i,J\!-\!1,k] & \\
     & \mid \x[i,J,k] = {\sf const} & \mbox{analogous} & \mathit{constant} \\
     & \mid \x[i,J,k]  = {\sf PHI}(\x_1[i,J,k],\x_2[i,J,k\!-\!1]) & & \mathit{pseudo} \; \mathit{PHI} \\
     & \mid \x[i,J,k] = {\sf raise\_dim}(\x'[i],(J\!:\!J,k\!:\!K)) & & \mathit{raise} \; \mathit{dimension(s)} \\
     & \mid \x[i,J] = {\sf drop\_dim}(\x'[i,J,k],k) &  & \mathit{drop} \; \mathit{dimension(s)} \\
%     & \mid \x = \y & & \mathit{propagation} \\
     & \mid {\sf for} \; i \; {\sf in} \; {\sf range}(I): s  & \gamma({\sf for} \; i \; {\sf in} \; {\sf range}(I): s) = & \mathit{loop} \\
     & & ~~\gamma(s)[0/i] \; ; \; \gamma(s)[1/i] \; ; \; ... \; ; \; \gamma(s)[I\!-\!1/i] & \\
  \end{array}    \\
\end{array}
\]
\vspace{-3mm}
\caption{MPC Source Syntax and Semantics. $\gamma$ defines the semantics of MPC source which is a linearization of MPC Source. A SIMD operation parallelizes operations across the vectorized $J$ dimension. $||$ denotes parallel execution, which is standard. $\gamma$ of a for loop unrolls the loop. $;$ denotes sequential execution. Iterative MPC Source trivially extends to non-vectorized dimensions over the enclosing loops.}\label{fig:MPC_Source_syntax}
\label{fig:mpc_syntax}
\vspace{-5mm}
\end{figure*}

@@

\figref{fig:mpc_syntax} states the syntax and linearization semantics of MPC Source. 
Although notation is heavy, the linearization simply produces schedules as discussed in~\secref{sec:model}; 
the iterative MPC Source gives rise to what we called sequential schedule where loops are unrolled 
and MPC Source with vectorized dimensions gives rise to what we called parallel schedule. 
For simplicity, we consider only scalars and read-only arrays, however, the treatment extends to write arrays as well.
$\x[i,J,k]$ denotes the value of scalar variable \x{} at loop nest $i,j,k$. 
%(We use 3 dimensions, $i,j,k$, however, the treatment generalizes to arbitrary dimensions.) 
Upper case $J$ denotes a vectorized dimension and lower case $i,k$ denote iterative dimensions. Our compiler imposes 
semantic restrictions over the syntax: (1) $\x$ is treated as a 3-dimensional array and (2) $\x[i,J,k]$ must be enclosed
into for-loops on non-vectorized dimensions $i$ and $k$:
\begin{pythonn}
for i in range(I):
   ...
   for k in range(K):
      ... x[i,J,k] ...
\end{pythonn}

@@

For each MPC Source program $a$ we compute the def-use edges in the standard way: if statement $s1 \in a$ defines variable \x, e.g., $\x[i,j,k] = ...$, 
and statement $s2 \in a$ uses \x, e.g., $... = ... \x[i,j,k]$ and there is a path in CFG from $s1$ to $s2$, then there is a def-use edge from
 $s1$ to $s2$. We extend the dimensionality of a statement into $s1[i,j,k]$ where $s1[i,j,k]$ inherits the dimensionality of the left-hand-side of the assignment.

@@

Array writes limit vectorization as they sometimes introduce infeasible loop-carried dependencies. Consider the classical example from~\cite{Aiken:1988}:
